Spike-and-Slab Generalized Additive Models and Scalable Algorithms for High-Dimensional Data

### 2021-09-21 Manuscript/02-Method.Rmd
Revise IWLS section

Similar to the EM-CD algorithm, the EM-IWLS algorithm is an iterative EM-based algorithm where iterative weighted least square algorithm are used to find the estimate of $\bs \beta, \phi$ that maximizes $E(Q_1)$, instead of the coordinate descent algorithm. It also can fit the models with a wider range of priors, e.g. mixture $t$ prior, including mixture normal prior as an special case, and mixture double exponential prior. The basic idea is to formulate the priors as hierarchical normal distributions and augment with the weighted normal likelihood of the generalized model, such that the model is feasible to fit via least squares algorithms. [@Yi2012] The primary purpose of the EM algorithm here is mitigate the computational burden associate with the hierarchical normal formulation.

A double exponential prior, $\beta|S \sim DE(0, S)$ can be formulated as a hierarchical normal prior with unknown variances, following gamma distributions, integrated out:
\begin{align*}
  \beta &\sim N(0, \tau^2)\\
  \tau^2 & \sim Gamma(1, 1/(2S^2)), 
\end{align*}
where $S = (1-\gamma)s_0 + \gamma s_1$ for spike-and-slab lasso priors. Now $Q_1$ needs to account for the hyper prior of $\tau^2$ and is expressed as 
\begin{equation}\label{eq:Q1_IWLS}
Q_1(\bs \beta, \phi) = \log f(\textbf{y}|\bs \beta, \phi) + \log f(\phi) + \sum\limits_{j=1}^p\left[\log f(\beta^0_j|{\tau^0}^2_j) + \log f({\tau^0}^2_j| S_j^0)+\sum\limits_{k=1}^{K_j} \{\log f(\beta^{\tp}_{jk}|{\tau^{\tp}}^2_{jk})+\log f({\tau^\tp}^2_{jk}| S^\tp_j)\}\right].
\end{equation}
As $\bs \tau^2$ are not of primary interest, we treat them as the "missing" data besides the latent indicators $\bs \gamma$, and hence need to construct the expectation $E_{\bs \gamma, \bs \tau^2|\Theta^{(t-1)}}(Q_1)$. To note, unlike the coefficients of the non-linear terms $\beta^\tp_{jk}$ which share the same latent indicator $\gamma^\tp_j$, $\tau^2_{jk}$ is coefficient specific for $\beta^\tp_{jk}$. Primarily, four quantities need to be calculate to formulate $E(Q_1), E({S_j^0}^{-1}|\beta_j^0, s_0, s_1), E(S^{-1}_j|\bs \beta_j^\tp, s_0, s_1), E({\tau^0}^2_{j}|S_j^0, \beta^0_j) \text{ and } E({\tau^\tp}^2_{jk}|S_j^\tp, \beta^\tp_{jk})$. As neither $E({S_j^0}^{-1}|\beta_j^0, s_0, s_1)$ nor $E(S^{-1}_j|\bs \beta_j^\tp, s_0, s_1)$ depends on $\tau^2$s, they can be derived via Bayes' theorem as in Equation (\ref{eq:exp_scale}). Shown in Yi and Ma [@Yi2012], the conditional posterior density of $\tau^{-2}$ is an inverse Gaussian distribution, and hence
\begin{align*}
 E({\tau^0}^{-2}_{j}|S_j^0, \beta^0_j) ={S_j^0}^{-1}/|\beta^0_j| \qquad E({\tau^\tp}^{-2}_{jk}|S_j^\tp, \beta^\tp_{jk})={S_j^\tp}^{-1}/|\beta^\tp_{jk}|,
\end{align*}
where $S_j^0$ and $S_j^\tp$ are replaced by the expectation.

To simplify $E(Q_1)$ by removing constant additive terms, we have 
\begin{equation}\label{eq:Q1_IWLS}
E(Q_1(\bs \beta, \phi)) = \log f(\textbf{y}|\bs \beta, \phi) + \log f(\phi) - \sum\limits_{j=1}^p\left[ \frac{({\beta_j^0}^2)E({\tau_j^0}^{-2}}{2\phi} +\sum\limits_{k=1}^{K_j} \frac{({\beta_{jk}^\tp}^2)E({\tau_{jk}^\tp}^{-2}}{2\phi}\right],
\end{equation}
where $\frac{({\beta_{jk}^\tp}^2)E({\tau_{jk}^\tp}^{-2}}{2\phi}$ is the kernel of a normal density with mean 0 and variance $E({\tau_{jk}^\tp}^{-2})\phi$.

Meanwhile, following the classical IWLS, we can approximate the generalized model likelihood at each iteration of the algorithms by the weighted normal likelihood:
$$
  f(\textbf{y}_i|\bs \beta, \phi) \approx Normal(\textbf{z}_i|X_i\beta, w_i^{-1}\phi)
$$
where the ‘normal response’ $z_i$ and ‘weight’ $w_i$ are called the pseudo-response and pseudo-weight respectively.The pseudo data and the pseudo-weight are calculated by
$$
\begin{aligned}
z_i &= \hat\eta_i - \frac{L^{'}(y_i|\hat\eta_i)}{L^{''}(y_i|\hat\eta_i)}& w_i &= - L^{''}(y_i|\hat\eta_i),
\end{aligned}
$$
where $L^{'}(y_i|\hat\eta_i, \hat \phi)$ and $L^{''}(y_i|\hat\eta_i, \hat \phi)$ are the first and second derivative of the log density, $\log f(\textbf{y}_i|\bs \beta, \phi)$. With $z\sim N(X\beta, \Sigma\phi)$ and $\beta \sim N(0, \phi\Sigma_{\tau^2})$, we can augment the two normal distributions and iterative solve for $\bs \beta$ and $\phi$.
  To note, as the EM-IWLS algorithm is based on the IWLS algorithm, an Hessian matrix can be derived. The Hessian matrix can be further used to construct variance-covariance matrix of coefficient estimates and in turn used for statistical inferences.

>>>>

Similar to the EM-CD algorithm, the EM-IWLS algorithm is an iterative EM-based algorithm where the iterative weighted least squares algorithm is used to find the estimate of $\bs \beta, \phi$ that maximizes $E(Q_1)$. The iterative weighted least squares algorithm was originally proposed to fit the classical generalized linear models, and generalized to fit some Bayesian hierarchical models.[@Gelman2013] Yi and Ma [@Yi2012] formulated Student's t-distribution and double exponential distribution as hierarchical normal distributions such that generalized linear models with shrinkage priors can be easily fitted using IWLS in combination with EM algorithm. In this work, we adapt the EM-IWLS paradigm to fit BHAM with spike-and-slab spline prior, with the full potential to adopt Student's t, double exponential and mixture Student's t priors when sparse assumption is not necessary.

<!-- The primary purpose of the EM algorithm here is mitigate the computational burden associate with the hierarchical normal formulation.  -->

A double exponential prior, $\beta|S \sim DE(0, S)$ can be formulated as a hierarchical normal prior with unknown variance $\tau^2 \sim Gamma(1, 1/(2S^2))$ integrated out:
\begin{align*}
  \beta &\sim N(0, \tau^2)\\
  \tau^2 & \sim Gamma(1, 1/(2S^2)), 
\end{align*}
For the mixture double exponential priors, we can define the scale parameter $S = (1-\gamma)s_0 + \gamma s_1$ following Equation (\ref{eq:ssl}). The change in the prior formulation in turns leads to the change in the log posterior density function, as $Q_1$ needs to account for the hyper prior of $\tau^2$: 
\begin{equation}\label{eq:Q1_IWLS}
Q_1(\bs \beta, \phi) = \log f(\textbf{y}|\bs \beta, \phi) + \log f(\phi) + \sum\limits_{j=1}^p\left[\log f(\beta^0_j|{\tau^0}^2_j) + \log f({\tau^0}^2_j| S_j^0)+\sum\limits_{k=1}^{K_j} \{\log f(\beta^{\tp}_{jk}|{\tau^{\tp}}^2_{jk})+\log f({\tau^\tp}^2_{jk}| S^\tp_j)\}\right].
\end{equation}
Since $\bs \tau^2$ are not of our primary interest, we treat them as the "missing" data in addition to the latent indicators $\bs \gamma$, and hence construct the expectation $E_{\bs \gamma, \bs \tau^2|\Theta^{(t-1)}}(Q_1)$ in the E-step. To note, unlike the same latent indicator $\gamma^\tp_j$ which is shared by the coefficients of the non-linear terms $\beta^\tp_{jk}$ for $k = 1, \dots, K_j$ , $\tau^2_{jk}$ is coefficient specific for $\beta^\tp_{jk}$. $E({S_j^0}^{-1}|\beta_j^0, s_0, s_1), E(S^{-1}_j|\bs \beta_j^\tp, s_0, s_1), E({\tau^0}^2_{j}|S_j^0, \beta^0_j) \text{ and } E({\tau^\tp}^2_{jk}|S_j^\tp, \beta^\tp_{jk})$ needs to be calculated to formulate $E(Q_1)$. As neither $E({S_j^0}^{-1}|\beta_j^0, s_0, s_1)$ nor $E(S^{-1}_j|\bs \beta_j^\tp, s_0, s_1)$ depends on $\tau^2$s, they can be derived using Equation (\ref{eq:exp_scale}) via Bayes' Theorem. On the other hand, $\tau^{2}$ following a gamma distribution is a conjugate prior for the normal variance, and the conditional posterior density of $\tau^{-2}$ is an inverse Gaussian distribution. $E({\tau^0}^{-2}_{j})$ and $E({\tau^\tp}^{-2}_{jk})$ are calculated using the closed form equation
\begin{align*}
 E({\tau^0}^{-2}_{j}|S_j^0, \beta^0_j) ={S_j^0}^{-1}/|\beta^0_j| \qquad E({\tau^\tp}^{-2}_{jk}|S_j^\tp, \beta^\tp_{jk})={S_j^\tp}^{-1}/|\beta^\tp_{jk}|,
\end{align*}
where $S_j^0$ and $S_j^\tp$ are replaced by the expectation and $\beta$s are replaced with $\beta^{(t-1)}$. With simplification (up to constant additive terms), we have 
\begin{equation}\label{eq:EQ1_IWLS}
E(Q_1) = \log f(\textbf{y}|\bs \beta, \phi) + \log f(\phi) - \sum\limits_{j=1}^p\left[ \frac{({\beta_j^0}^2)}{2\phi E{\tau_j^0}^{2}} +\sum\limits_{k=1}^{K_j} \frac{{\beta_{jk}^\tp}^2}{2\phi E({\tau_{jk}^\tp}^{2})}\right].
\end{equation}
$\frac{\beta^2}{2\phi E(\tau^{2})}$ can be seen as the kernel of a normal density with mean 0 and variance $\phi E(\tau^{2})$.

Meanwhile, following the classical IWLS, we can approximate the generalized model likelihood at each iteration with a weighted normal likelihood:
$$
  f(\textbf{y}|\bs \beta, \phi) \approx MVN(\textbf{z}|X\beta, \Sigma w_i^{-1}\phi)
$$
where the ‘normal response’ $z_i$ and ‘weight’ $w_i$ are called the pseudo-response and pseudo-weight respectively.The pseudo data and the pseudo-weight are calculated by
$$
\begin{aligned}
z_i &= \hat\eta_i - \frac{L^{'}(y_i|\hat\eta_i)}{L^{''}(y_i|\hat\eta_i)}& w_i &= - L^{''}(y_i|\hat\eta_i),
\end{aligned}
$$
where $L^{'}(y_i|\hat\eta_i, \hat \phi)$ and $L^{''}(y_i|\hat\eta_i, \hat \phi)$ are the first and second derivative of the log density, $\log f(\textbf{y}_i|\bs \beta, \phi)$. With $z\sim N(X\beta, \Sigma\phi)$ and $\beta \sim N(0, \phi\Sigma_{\tau^2})$, we can augment the two normal distributions and iterative solve for $\bs \beta$ and $\phi$.
  To note, as the EM-IWLS algorithm is based on the IWLS algorithm, an Hessian matrix can be derived. The Hessian matrix can be further used to construct variance-covariance matrix of coefficient estimates and in turn used for statistical inferences.
